---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
*Load Required Packages*
```{r}
if (!require("lattice")) {install.packages("lattice")}
require(ggplot2)

if (!require("plyr")) {install.packages("plyr")}
require(plyr)

if (!require("knitr")) {install.packages("knitr")}
require(knitr)
```
*Download data if necessary*
```{r}
if (!file.exists("./activity_data.zip")) {
        fileURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileURL, "./activity_data.zip", method = "curl")
        unzip("./activity_data.zip", overwrite = T)
        dateDownloaded <- date()
        dateDownloaded
}
```
*Read and Transform Data*  
Data is read and the structure of the data frame is displayed.
```{r}
activity<-read.csv("activity.csv", colClasses=c("integer","Date","integer"))
str(activity)
```
A summary of the data is also provided.  As we can see, there are many null values in the steps data.
```{r}
summary(activity)
```
## What is mean total number of steps taken per day?
```{r}
totalSteps<-tapply(activity$steps,activity$date,sum,na.rm=T)
stepsMean<-mean(totalSteps)
stepsMean
```
```{r}
stepsMedian<-median(totalSteps)
stepsMedian
```
*Histogram of Total Steps/ Day*
```{r}
hist(totalSteps, breaks=10, main="Histogram of Total Steps per Day",xlab="Total Steps/ Day")
abline(v=stepsMean,col="Red", lwd=2)
abline(v=stepsMedian,col="Green", lwd=2)
legend(x="topright", legend=c("Mean","Median"), col=c("Red","Green"),lwd=2)
```


## What is the average daily activity pattern?
*Calculate the average across all days and plot*  
  - Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
avgSteps <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)
plot(steps ~ interval, data = avgSteps, main="Avg Steps per Day", ylab="Avg Steps", type = "l")
```
  
  - Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
  *The interval with the maximum number of average steps is calculated below*
```{r}
maxInterval<-avgSteps[which.max(avgSteps$steps), ]$interval
maxInterval
```

## Imputing missing values  
  - Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
sum(is.na(activity))
```
  - Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
  
*The mean can be used to approximate the missing values.  A summary is provided to demonstrate that the null values have been removed.*
```{r}
activity_nonull <- activity
activity_nonull$steps[is.na(activity_nonull$steps)] <- mean(activity_nonull$steps, na.rm = T)
summary(activity_nonull)
```
  - Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
totalSteps_nonull<-tapply(activity_nonull$steps,activity_nonull$date,sum,na.rm=T)
stepsMean_nonull<-mean(totalSteps_nonull)
stepsMean_nonull
```
```{r}
stepsMedian_nonull<-median(totalSteps_nonull)
stepsMedian_nonull
```
*A side-by-side plot of the two datasets is provided for comparison.*
```{r}
par(mar=c(2,2,2,2),mfcol=c(1,2))
hist(totalSteps, breaks=10, main="TTL Steps/Day with null values",xlab="Total Steps/ Day")
abline(v=stepsMean,col="Red", lwd=2)
abline(v=stepsMedian,col="Green", lwd=2)
hist(totalSteps_nonull, breaks=10, main="TTL Steps/Day without null values",xlab="Total Steps/ Day")
abline(v=stepsMean_nonull,col="Red", lwd=2)
abline(v=stepsMedian_nonull,col="Green", lwd=2)
```
  
  - Do these values differ from the estimates from the first part of the assignment?  
        *Yes. The average steps per day vary in each interval that originally had null values.*  
  - What is the impact of imputing missing data on the estimates of the total daily number of steps?  
        *The results of the analysis will change based on the method used to impute the missing data*  

## Are there differences in activity patterns between weekdays and weekends?  
  - For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.  
  
  - Create a new factor variable in the dataset with two levels  weekday and weekend indicating whether a given date is a weekday or weekend day.
  
```{r}
activity_nonull$dayCat<-factor(weekdays(activity_nonull$date) %in% c("Saturday","Sunday"),
                               labels=c("weekday","weekend"), ordered=FALSE)
summary(activity_nonull)
```
  
  - Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r}
avgSteps_nonull <- aggregate(steps ~ interval + dayCat, data = activity_nonull, mean)
xyplot(steps ~ interval | dayCat, avgSteps_nonull, type = "l", layout = c(1, 2), 
       xlab = "Interval", ylab = "Avg Number of Steps", 
       main="Comparison of Avg Steps by Weekend vs. Weekday")
```
  
*The charts demonstrate that activity continues at a higher rate thoughout the day on Weekends*